---
title: "Introduction to fastDiversity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to fastDiversity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 6,
  fig.height = 4
)
```

## Overview

`fastDiversity` provides fast and flexible calculation of standard population
genetic diversity statistics from biallelic SNP data. It is designed for
datasets where samples span multiple populations, sites, or species that need
to be analysed simultaneously, a common scenario in conservation genomics and
comparative population studies.

The package accepts genotype data in standard biallelic dosage format: `0`
(homozygous reference, *aa*), `1` (heterozygous, *aA*), and `2` (homozygous
alternate, *AA*), with missing data as `NA`. This format is the output produced by most SNP-calling pipelines.

---

## Installation

```{r install, eval = FALSE}
remotes::install_github("eilishmcmaster/fastDiversity")
```

```{r load, message = FALSE, warning = FALSE}
library(fastDiversity)
library(dplyr)
library(ggplot2)
library(ggthemes)
```

---

## Example dataset: *Doryphora sassafras*

The package ships with SNP data from *Doryphora sassafras* (Sassafras,
Atherospermataceae), a rainforest tree endemic to the eastern Australian coast.
The dataset spans five genetic populations, each sampled across multiple sites.
This structure, multiple hierarchical grouping levels within a single species,
is the typical use case for `fastDiversity`.

```{r load-data}
data(example_data)

# Genotype matrix: rows are samples, columns are SNP loci
# Values are 0 (aa), 1 (aA), 2 (AA), or NA (missing)
dim(example_gt)

example_print <- example_gt[c(1,100,150,200,250), 1:10]
colnames(example_print) <- NULL
example_print

```

```{r inspect-meta}
# Metadata: one row per sample, matching the row order of example_gt
head(example_meta)

# Five genetic populations, each with multiple sampling sites
table(example_meta$site, example_meta$population)
```

The row order of `example_gt` must match the row order of `example_meta`,
the nth row of the genotype matrix corresponds to the nth row of the metadata.

--- 

## Principal coordinates analysis

A distance-based PCA provides an intuitive overview of genetic structure across
individuals. PC scores are computed from pairwise genetic distances and merged
with metadata for plotting:

```{r pca, fig.width = 6.5, fig.height = 5}
pca   <- prcomp(dist(example_gt), center = TRUE, scale. = TRUE)
s_pca <- summary(pca)

# Axis labels showing % variance explained
importance <- s_pca[["importance"]][2, ]
axis_names  <- paste0(names(importance), " (", round(importance * 100, 1), "%)")

# Merge PC scores with sample metadata
pcs <- pca$x %>%
  data.frame() %>%
  merge(example_meta, by.x = 0, by.y = "sample")

ggplot(pcs, aes(x = PC1, y = PC2, color = factor(population))) +
  geom_point(size = 2, alpha = 0.8) +
  labs(
    title  = expression("PCA of SNP genotypes"),
    x      = axis_names[1],
    y      = axis_names[2],
    color  = "Population"
  ) +
  theme_bw()
```

Clustering of samples in PCA space corroborates the population structure
indicated by the diversity statistics and private allele analysis.


---

## Allele sharing and private alleles

Understanding which alleles are unique to a single population (private) versus
shared across populations is a useful first step in assessing population
distinctiveness. Populations with many private alleles may represent distinct
evolutionary lineages that warrant separate management.

### Step 1: Build the allele list

`make_allele_list()` takes the genotype matrix and a grouping vector and
returns a named list of allele sets, one per group.

```{r allele-list}
allele_list_population <- make_allele_list(example_gt, example_meta$population)

names(allele_list_population)
```

### Step 2: Visualise allele overlap

Because the result is a standard named list, it works directly with `ggvenn`:

```{r venn, fig.width = 5.5, fig.height = 5.5, message = FALSE}
ggvenn::ggvenn(allele_list_population, 
               stroke_size=0)+
  ggtitle("Allele sharing among populations")
```

Alternatively, an upset plot may be easier to read: 

```{r venn upset plot, fig.width = 8, fig.height = 6, message = FALSE}
ggVennDiagram::plot_upset(ggVennDiagram::Venn(allele_list_population), sets.bar.show.numbers = FALSE)
```

This shows how much allelic diversity is shared across populations versus
restricted to individual groups. A large exclusive region indicates a
genetically distinct population.

### Step 3: Count private and total alleles

```{r private-alleles}
private_total_alleles_population <- calculate_private_alleles(allele_list_population)

private_total_alleles_population
```

The output gives private allele counts (alleles found in only one population)
and total allele counts per group. These can be used directly in reporting or
joined to the main statistics table (shown below).

---

## Diversity statistics with `faststats()`

`faststats()` calculates allelic richness, heterozygosity, inbreeding
coefficients, and population differentiation for each combination of genetic
group and site, with locus-level filtering applied independently within each
genetic group.

### Filtering parameters

| Parameter | Description |
|-----------|-------------|
| `minimum_n`| Exclude sites with fewer than this many individuals |
| `minimum_loci`| Exclude genetic groups where fewer than this many loci pass filtering |
| `maf`| Minor allele frequency threshold: loci below this value are excluded per genetic group (typically 0.05) |
| `max_missingness`| Exclude loci missing in more than this proportion of samples within a group (e.g. 0.3 removes loci missing in >30% of samples) |

Filtering is applied within each genetic group independently. A locus retained
for one population may be excluded for another if it fails MAF or missingness
thresholds in that group, an important distinction from global pre-filtering,
which can retain uninformative loci in some groups or discard informative ones
in others.

### Running faststats

`genetic_group_variable` defines the level at which locus filtering is applied
(e.g. species or population). `site_variable` defines the grouping unit for
which statistics are calculated (e.g. sampling site). The function returns one
row per unique genetic group × site combination.

```{r faststats}
basicstats <- faststats(
  gt                    = example_gt,
  genetic_group_variable = example_meta$population,
  site_variable         = example_meta$site,
  minimum_n             = 6,
  minimum_loci          = 50,
  maf                   = 0.05,
  max_missingness       = 0.3
)

head(basicstats)
```

### Output columns explained

| Column | Description |
|--------|-------------|
| `group` | Genetic group ID (e.g. population, species) |
| `site` | Site ID |
| `Ar` | Allelic richness per site per group, standardised for sample size |
| `Ho` | Observed heterozygosity: proportion of heterozygous loci averaged across individuals |
| `He` | Expected heterozygosity under Hardy-Weinberg equilibrium within site_variable |
| `uHe` | Unbiased expected heterozygosity (corrected for small sample size) |
| `Fis` | Inbreeding coefficient (1 - Ho/He). Positive values indicate excess homozygosity; negative values indicate excess heterozygosity |
| `uFis` | Unbiased inbreeding coefficient (1 - Ho/uHe) |
| `Fst` | Population differentiation: Weir & Cockerham (1984) FST, calculated as (Ht - Hs) / Ht |
| `loci` | Number of loci retained after filtering for this group |
| `n` | Number of individuals sampled at this site x group combination |


### Plotting diversity statistics

Because `faststats()` returns one row per site, each population contributes
multiple points, one per site. A faceted multi-panel plot gives a
simultaneous overview of heterozygosity, inbreeding, and differentiation:

```{r plot-multi, fig.width = 6, fig.height = 4, message = FALSE, warning = FALSE}
library(tidyr)

stats_long <- basicstats %>%
  mutate(across(c(Ho, He, Fis, Fst), as.numeric)) %>%
  pivot_longer(
    cols      = c(Ho, He, Fis, Fst),
    names_to  = "statistic",
    values_to = "value"
  ) %>%
  mutate(statistic = factor(statistic,
    levels = c("Ho", "He", "Fis", "Fst"),
    labels = c("Observed heterozygosity (Ho)",
               "Expected heterozygosity (He)",
               "Inbreeding coefficient (Fis)",
               "Fixation index (Fst)")
  ))

ggplot(stats_long, aes(x = group, y = value, fill = group)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.7) +
  facet_wrap(~ statistic, scales = "free_y", ncol = 2) +
  labs(
    title = "Diversity statistics",
    x     = "Population",
    y     = NULL
  ) +
  theme_bw() +
  theme(
    legend.position  = "none",
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(size = 9)
  )
```

### Optional: bootstrapped confidence intervals

For published reporting, bootstrapped confidence intervals can be added around
Ho, He, and Fis. Two resampling strategies are available:

- **By loci** (`get_CI_loci_resampling = TRUE`): analogous to
  `hierfstat::boot.ppfis` : loci are resampled with replacement to quantify
  how sensitive statistics are to the choice of markers.
- **By individuals** (`get_CI_individual_resampling = TRUE`): analogous to
  `diveRsity::basicstats` : individuals are resampled with replacement to
  quantify sampling uncertainty.

```{r faststats-ci, eval = FALSE}
basicstats_ci <- faststats(
  gt                     = example_gt,
  genetic_group_variable = example_meta$population,
  site_variable          = example_meta$site,
  minimum_n              = 6,
  minimum_loci           = 50,
  maf                    = 0.05,
  max_missingness        = 0.3,
  get_CI_loci_resampling = TRUE,
  boots                  = 100,
  CI_alpha               = 0.05
)
```

CI columns are appended with a prefix for the resampling strategy (`loci_` or
`ind_`) and a suffix for the statistic and quantile boundary (e.g.
`loci_ho.2.5%`, `loci_ho.97.5%`). These plug directly into
`geom_pointrange()`:

```{r plot-ci, eval = FALSE}
ggplot(basicstats_ci, aes(x = group, y = as.numeric(Ho), colour = group)) +
  geom_pointrange(
    aes(
      ymin = as.numeric(`loci_ho.2.5%`),
      ymax = as.numeric(`loci_ho.97.5%`)
    ),
    position = position_jitter(width = 0.2, seed = 1),
    size = 0.6
  ) +
  labs(
    title = expression("Ho ± 95% CI (loci bootstrap)"),
    x = "Population",
    y = "Observed heterozygosity (Ho)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

### Optional: Hardy-Weinberg equilibrium testing

`faststats()` can test whether allele frequencies at each site deviate
significantly from Hardy-Weinberg equilibrium expectations using Fisher's
exact test, with p-values combined across loci using Fisher's method:

```{r faststats-hwe, eval = FALSE}
basicstats_hwe <- suppressWarnings( faststats(
  gt                     = example_gt,
  genetic_group_variable = example_meta$population,
  site_variable          = example_meta$population,
  minimum_n              = 6,
  minimum_loci           = 50,
  maf                    = 0.05,
  max_missingness        = 0.3,
  run_HWE_test           = TRUE,
  return_locus_stats     = FALSE   # set TRUE to retrieve per-locus p-values
))

basicstats_hwe[,c("site", "Ho", "He", "Fis", "Fst", "loci", "n", "HWE_p")]
```

The `HW_fisher_p` column gives the combined p-value per site. Sites with
values below the chosen alpha threshold show significant deviation from HWE,
which may reflect inbreeding, cryptic population structure within a site,
selection at linked loci, or genotyping artefacts.


---
## Individual-level statistics

In addition to site-level summaries, `fastDiversity` provides two functions
for calculating statistics at the level of individual samples. These are useful
for examining variation in inbreeding and heterozygosity on an individual basis,
which can be informative for fitness and hybridisation.

Both functions apply the same locus-level filtering as `faststats()`: MAF and
missingness thresholds are applied per genetic group before calculation, ensuring
that individual estimates are based on the same high-quality, informative loci
used in population-level analyses.

### Individual multilocus heterozygosity

`individual_Ho()` calculates multilocus heterozygosity (MLH) for each sample:
the proportion of retained loci that are heterozygous in that individual.

```{r ind-ho}
ind_MLH_ho <- individual_Ho(
  example_gt,
  genetic_group_variable = example_meta$population,
  maf                    = 0.05,
  max_missingness        = 0.3
)

head(ind_MLH_ho)
```

### Individual inbreeding coefficient (Wright's *F*)

`individual_F()` calculates Wright's inbreeding coefficient for each sample:
*F* = 1 − (Ho / He), where He is calculated across the genetic group. This
measures the probability that both alleles at a locus are identical by descent
(Frankham, Ballou & Briscoe 2010). Because He is calculated at the group level
rather than across genetic groups, estimates are not deflated by the Wahlund effect
that would arise from pooling genetically differentiated individuals.

```{r ind-f}
ind_F <- individual_F(
  example_gt,
  genetic_group_variable = example_meta$population,
  maf                    = 0.05,
  max_missingness        = 0.3
)

head(ind_F)
```

### Relating MLH to *F*

Plotting individual MLH against *F* reveals the expected negative relationship:
individuals with higher observed heterozygosity have lower inbreeding
coefficients. 

```{r plot-ind, fig.width = 5.5, fig.height = 4.5}
# Combine into a data frame and add population labels
ind_df <- data.frame(
  sample     = names(ind_MLH_ho),
  MLH        = as.numeric(ind_MLH_ho),
  F          = as.numeric(ind_F)
) %>%
  merge(example_meta, by = "sample")

ggplot(ind_df, aes(x = MLH, y = F, colour = as.factor(population))) +
  geom_point(size = 2, alpha = 0.8) +
  labs(
    title    = expression("Individual MLH vs inbreeding coefficient"),
    x        = "Multilocus heterozygosity (Ho)",
    y        = expression(paste("Inbreeding coefficient (", italic(F), ")")),
    colour   = "Population"
  ) +
  theme_bw()
```


## Nucleotide diversity and Tajima's D

`pi_theta_d_stats()` calculates nucleotide diversity (pi), Wu & Watterson's theta
(W), the number of segregating sites (S), and Tajima's D for each genetic
site_variable specified within genetic_group_variable. The implementation follows the missing-data-aware
approach of Korunes & Samuk (2021), meaning that loci with missing genotypes
contribute proportionally rather than being excluded entirely, an important
correction for RADseq and DArTseq data where missingness is common.

### Running pi_theta_d_stats

When `n = NULL`, all individuals at each site are used. The `n` argument can
be set to a fixed integer to subsample individuals, which allows comparison
across sites with unequal sample sizes:

```{r tajima-run}
# Use all available individuals per population
D_n_all <- pi_theta_d_stats(
  example_gt,
  genetic_group_variable = example_meta$population,
  site_variable          = example_meta$population,
  n                      = NULL,
  minimum_n              = 5,
  minimum_loci           = 5,
  max_missingness        = 0.1,
  maf                    = 0,
  allele_count_min       = 0
)

# Subsample to 20 individuals per population
D_n_20 <- pi_theta_d_stats(
  example_gt,
  genetic_group_variable = example_meta$population,
  site_variable          = example_meta$population,
  n                      = 20,
  minimum_n              = 5,
  minimum_loci           = 5,
  max_missingness        = 0.1,
  maf                    = 0,
  allele_count_min       = 0
)

# Subsample to 10 individuals per population
D_n_10 <- pi_theta_d_stats(
  example_gt,
  genetic_group_variable = example_meta$population,
  site_variable          = example_meta$population,
  n                      = 10,
  minimum_n              = 5,
  minimum_loci           = 5,
  max_missingness        = 0.1,
  maf                    = 0,
  allele_count_min       = 0
)

# Subsample to 5 individuals per population
D_n_5 <- pi_theta_d_stats(
  example_gt,
  genetic_group_variable = example_meta$population,
  site_variable          = example_meta$population,
  n                      = 5,
  minimum_n              = 5,
  minimum_loci           = 5,
  max_missingness        = 0.1,
  maf                    = 0,
  allele_count_min       = 0
)
```

### Output columns explained

| Column | Description |
|--------|-------------|
| `group` | Genetic group ID |
| `site` | Site ID |
| `loci` | Number of loci retained after filtering |
| `n` | Number of individuals used |
| `pi` | Nucleotide diversity: mean pairwise differences across loci |
| `S` | Number of segregating (polymorphic) sites |
| `theta` | Wu & Watterson's theta: diversity estimate based on S, corrected for sample size |
| `D` | Tajima's D: standardised difference between pi and W. Negative values suggest an excess of rare variants (population expansion or purifying selection); positive values suggest an excess of intermediate-frequency variants (balancing selection or population contraction) |
| `real_missingness` | Proportion of missing data in the filtered matrix used for calculation |

### Effect of sample size on diversity estimates

A key feature of `pi_theta_d_stats()` is the `n` argument, which subsamples
individuals to a fixed number before calculation. This allows fair comparison
across populations with unequal sample sizes. However, sample size affects
estimates differently: S is particularly sensitive to n because larger samples
are more likely to capture rare variants, which inflates W relative to pi and
therefore pulls D negative.

The plot below illustrates this by comparing estimates across three levels of
subsampling:

```{r tajima-plot, fig.width = 6, fig.height = 6, message = FALSE, warning = FALSE}
library(tidyr)

# Combine results and label by sample size used
D_combined <- bind_rows(
  D_n_all %>% mutate(sample_size = paste0("All (n=", n, ")")),
  D_n_20  %>% mutate(sample_size = "n = 20"),
  D_n_10  %>% mutate(sample_size = "n = 10"),
  D_n_5   %>% mutate(sample_size = "n = 5")
) %>%
  mutate(
    across(c(pi, S, theta, D), as.numeric),
    sample_size = factor(sample_size,
                         levels = c("n = 5", "n = 10", "n = 20",
                                    unique(D_n_all %>%
                                             mutate(l = paste0("All (n=", n, ")")) %>%
                                             pull(l))))
  )

# Reshape to long for faceting
D_long <- D_combined %>%
  pivot_longer(cols = c(pi, S, theta, D),
               names_to  = "statistic",
               values_to = "value") %>%
  mutate(statistic = factor(statistic,
    levels = c("pi", "theta", "S", "D"),
    labels = c("Nucleotide diversity (pi)",
               "Wu & Watterson's theta (W)",
               "Segregating sites (S)",
               "Tajima's D")))

ggplot(D_long, aes(x = group, y = value, colour = sample_size,
                   group = sample_size)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ statistic, scales = "free_y", ncol = 2) +
  labs(
    title   = expression("Effect of sample size on diversity statistics"),
    x       = "Population",
    y       = NULL,
    colour  = "Sample size"
  ) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "grey92"),
    strip.text       = element_text(size = 9),
    legend.position  = "bottom"
  )
```



## References

Goudet, J. (2005). Hierfstat, a package for r to compute and test hierarchical F-statistics. 
*Molecular Ecology Notes*, 5(1), 184–186. https://doi.org/10.1111/j.1471-8286.2004.00828.x

Kalinowski, S. T. (2004). hp-rare 1.0: a computer program for performing
rarefaction on measures of allelic richness. *Molecular Ecology Notes*, 5(1),
187–189. https://doi.org/10.1111/j.1471-8286.2004.00845.x

Keenan, K. G., McGinnity, P., Cross, T. F., Crozier, W. W., & Prodöhl, P. A.
(2013). diveRsity: An R package for the estimation and exploration of
population genetics parameters and their associated errors. *Methods in
Ecology and Evolution*, 4(8), 782–788.
https://doi.org/10.1111/2041-210x.12067

Korunes, K. L., & Samuk, K. (2021). pixy: Unbiased estimation of nucleotide diversity 
and divergence in the presence of missing data. *Molecular Ecology Resources*,
21(4), 1359–1368. https://doi.org/10.1111/1755-0998.13326

Nei, M. (1978). Estimation of average heterozygosity and genetic distance from
a small number of individuals. *Genetics*, 89(3), 583–590.
https://doi.org/10.1093/genetics/89.3.583
